$\XX$ dubious ways of estimating $\rr$

### 2018-07-03 paper/Version9/draft_v9.tex
finally on git

\textit{We compare the different ways to estimate $\rr$ to see which ones give more accurate and reliable estimates.}  As there is no way to know the ``true'' $\rr$ of an actual disease we compare our methods by fitting them to simulated data sets and look at the resulting $\rr$ estimates and standard errors. Our main comparisons are done with data generated from the SIR model with random error added so that we can compare model estimates to a known true $\rr$ value. We vary the conditions under which the SIR data is generated to see how the results from the estimators change. We also look at data from different functional models to see how the estimators perform under different circumstances.

@@

 The data from the linear SIR model were simulated with parameters $\beta$ = 0.06 and $\gamma$ = 0.05, such that $\rr$ = 1.2. This model will allow us to compare how the SIR models perform under misspecified differential equations.

 \textbf{Random error is added to the data generated here as well using the same 4 error models as above. The other specifications for these datasets are the same as for the baseline classic SIR data, namely the variance of the added errors and starting compartment sizes.}

@@

Standard errors for the maximum likelihood and branching process models are calculated using the negative reciprocal of the Fisher information. The sequential Bayes model uses the standard deviation from the posterior distribution. We use the jackknife method to estimate the standard error of the SIR smoothed max and the SIR smoothed IPR methods. The jackknife calculates the standard error using the standard deviation of estimates of $\rr$ after leaving out a single data point.

Some of the models use only one single point in time to estimate $\rr$ and therefore do not seem to have an innate source of variance. We currently do not have standard errors for the SIR Max and SIR IPR models.

@@

\textbf{STILL NEED TO ACTUALLY DO DPE}

We do not have a direct way to estimate $/rr$ using the survival function method from the data we have. The method requires a density estimate of the survival function at each point in time. Either we would need to have some prior knowledge or assumptions about what the survival function for a particular disease was, or we would need to search over a space of possible function and possible parameter values to estimate $/rr$. For similar reasons, we can not estimate $/rr$ using the Von Forester equations using the data we have, as we would need to know the infectivity of a diseased individual over time. This may be possible for a known, well studied disease, but may not be  For the direct parameter estimation method, assuming the population is homogeneous and that we know the average duration of infection, we can estimate the probability of transmission parameter and its standard error by averaging over the ratio between percent change in the susceptible compartment and the prevalence of the disease at each time step.

@@

\paragraph{SIR Models}
%In general, the models derived from the SIR ordinary differential equations are relatively accurate. Most estimates are within the range of $\pm$ 0.2 of the true $\rr$ for the baseline SIR datasets. The magnitude of the standard errors on the baseline datasets for the different SIR models generally depend on the method used. The models that estimate $\rr$ using optimization or gradient descent, namely SIR LS and SIR ReLS, have standard errors on the order of $10^{-4}$ to $10^{-7}$. The standard errors are somewhat larger for the methods that use regression, on the order of $10^{-1}$ to $10^{-3}$.

%The SIR models as a whole also do fairly well with changing the starting number of infected people. However, they tend not to do as well when the variance of the added errors is very large; in these situations, many of the models tend to have very large standard errors, sometimes on the order of $10^{1}$ or $10^{2}$.

%The exceptions to the above are the linear, incidence-prevalence ratio, and Markov chain models. The estimates tend to be far from the true $\rr$, and they fluctuate greatly between the baseline data sets. The standard errors for these models (if they exist) also tend to be very large under some of the changing conditions.

%Many of the SIR models do not work as well with the data that come from the polynomial linear models. Specifically, for the 0th and 1st order model datasets, the estimates can be unreasonably large or negative. The Log linear, Sequential Bayes, Smoothed max, and Max models do not have this problem.

@@

%All of these models use optimization to estimate the parameters that are used to calculate $\rr$. They tend to be fairly accurate and consistent on the baseline data, often as good as or better than those from the various SIR models. The SEIR models (least squares and next generation) do not handle the large added errors very well, as it tends to give ridiculously large results. The standard errors tend to be on the order of $10^{-2}$ to $10^{-3}$. However, as with the estimates above, the standard errors from the SEIR models can become unreasonably large when the variance of the added errors becomes large. Also, as with the SIR models, the $\rr$ estimates and standard errors from the 0th and 1st order polynomial linear model datasets are not very good, as the former is sometimes negative and the latter can be very large. The MSEIR model tends to do fairly well in all situations.

@@

%We have assumed for these models that the number of infections follows an exponential curve throughout the entire time frame of the data sets. This assumption certainly does not hold for the data from the 0th and 1st order linear models, and may not hold for some of the other data sets either.

%These models are generally not as accurate as the compartment models above. Most of them, except for the initial size model, tend to underestimate $\rr$, with estimates between 0.8 and 1.0. The standard errors are all fairly small, on the order of magnitude from $10^{-1}$ to $10^{-3}$. This is larger than many of the compartment model standard errors. The main advantage that all of these models, except for the maximum likelihood model, seem to be more robust to changing conditions and functional forms of the data compared to the compartment models. The estimates and standard errors for $\rr$ appear to be reasonable in all cases, including for very large added errors and for the 0th and 1st order linear model data.

@@

%The branching process model (the only one in this category so far) is not very accurate. In particular, for the SIR datasets, the $\rr$ estimates are less than 1, while the true $\rr$ is greater than 1. The standard errors, though, are relatively small, around the same order of magnitude as the SIR least squares models. Model estimates and standard errors are fairly consistent across the baseline datasets and changing starting compartment sizes. However, increasing the standard deviation of the added errors tends to have a somewhat noticeable impact on model results. The model also does not handle the 0th order polynomial linear data well.

@@

Below, we compare the estimates and standard errors of $\rr$ given by each of the models for a few selected data sets using forest plots; they show the point estimates and 95\% confidence intervals of $\rr$ from each model for a particular data set. The true $\rr$ value is also shown by the vertical red line in the graphs. All of the figures below show results from data sets that were simulated with added Gaussian errors. The plot on the top left shows the estimates from one of the baseline data sets. For the most part, the $\rr$ values range from approximately 0.8 to 2. We note that many of the models estimate an $/rr$ value of approximately 1, and some estimate that $/rr$ is less than 1. This would indicate that the infection disease is growing very slowly or is actually dying out, when in reality the disease is spreading at a fairly rapid rate. The confidence intervals are very small relative to the estimates for the most part; exceptions to this include the SIR IPR and LMA models as well as the Markov chain model.

The plot on the top right shows the estimates for the data set where $Y_0$ is increased to 1000. The $\rr$ estimates are almost all further away from the true $\rr$ value compared to the results from the baseline data sets. The confidence intervals are also more often than not larger than those in the baseline. In particular, the standard errors for some of the SIR models increase by several orders of magnitude.

The bottom left plot shows the estimates for the data set where the error added to the generated data is increased. For most of the models, the estimates of $\rr$ decrease slightly, and overall they do not change very much compared to the baseline data set. The confidence intervals for this data set increase substantially for some models, such as the SIR LMA, while they remain approximately the same as the baseline data sets for several of the others.

On the bottom right, we see the $\rr$ estimates for the case when disease prevalence grows linearly, as opposed to exponentially as seen in the data sets generated under SIR conditions, where the disease prevalence grows exponentially. Model performance here varies depending on how well the models the aforementioned linear growth. Some of the models, such as the SIR Markov chain and the IPR models, do not handle this case very well, giving much larger estimates and standard errors than are reasonable. On the other hand, models such as the SIR maximum and final size seem to be more flexible to different trajectories.

>>>>

\textit{With the many different ways to estimate $\rr$, we want to compare them all to see which ones give more accurate and reliable estimates.}  As there is no way to know the ``true'' $\rr$ of an actual disease we compare our methods by fitting them to simulated data sets and look at the resulting $\rr$ estimates and standard errors. Our main comparisons are done with data generated from the SIR model with random error added so that we can compare model estimates to a known true $\rr$ value. We vary the conditions under which the SIR data is generated to see how the results from the estimators change. We also look at data from different functional models to see how the estimators perform under different circumstances.

@@

 The data from the linear SIR model were simulated with parameters $\beta$ = 0.06 and $\gamma$ = 0.05, such that $\rr$ = 1.2.

 \textbf{Random error is added to the data generated here as well, using the same 4 error models as above. The other specifications for these datasets are the same as for the baseline classic SIR data, namely the variance of the added errors and starting compartment sizes.}

@@

\textbf{Standard errors for the maximum likelihood and branching process models are calculated using the negative reciprocal of the Fisher information. The sequential Bayes model uses the standard deviation from the posterior distribution. We use the jackknife method to estimate the standard error of the SIR smoothed max and the SIR smoothed IPR methods. The jackknife calculates the standard error using the standard deviation of estimates of $\rr$ after leaving out a single data point.}

\textbf{Some of the models use only one single point in time to estimate $\rr$ and therefore do not seem to have an innate source of variance. We currently do not have standard errors for the SIR Max and SIR IPR models.}

@@

%% How do we want to estimate this?????????
%% SG:
%% 3.1.1:  So I think this one we should mention there is no direct way to estimate this from the simple data we have.  Alternatively, you could do a parameter sweep with different beta distributions for F(a) and beta for b(a).
%% 3.1.2:  So assuming the population is homogeneous, k = N, D = gamma^{-1}.  So we have to estimmate $b$ from the data.  So I would average over (change in % susceptible ) / old # infected and use that as $b$.
%% 3.1.3:  No clue.

@@

\paragraph{SIR Models}
In general, the models derived from the SIR ordinary differential equations are relatively accurate. Most estimates are within the range of $\pm$ 0.2 of the true $\rr$ for the baseline SIR datasets. The magnitude of the standard errors on the baseline datasets for the different SIR models generally depend on the method used. The models that estimate $\rr$ using optimization or gradient descent, namely SIR LS and SIR ReLS, have standard errors on the order of $10^{-4}$ to $10^{-7}$. The standard errors are somewhat larger for the methods that use regression, on the order of $10^{-1}$ to $10^{-3}$.

The SIR models as a whole also do fairly well with changing the starting number of infected people. However, they tend not to do as well when the variance of the added errors is very large; in these situations, many of the models tend to have very large standard errors, sometimes on the order of $10^{1}$ or $10^{2}$.

The exceptions to the above are the linear, incidence-prevalence ratio, and Markov chain models. The estimates tend to be far from the true $\rr$, and they fluctuate greatly between the baseline data sets. The standard errors for these models (if they exist) also tend to be very large under some of the changing conditions.

Many of the SIR models do not work as well with the data that come from the polynomial linear models. Specifically, for the 0th and 1st order model datasets, the estimates can be unreasonably large or negative. The Log linear, Sequential Bayes, Smoothed max, and Max models do not have this problem.

@@

All of these models use optimization to estimate the parameters that are used to calculate $\rr$. They tend to be fairly accurate and consistent on the baseline data, often as good as or better than those from the various SIR models. The SEIR models (least squares and next generation) do not handle the large added errors very well, as it tends to give ridiculously large results. The standard errors tend to be on the order of $10^{-2}$ to $10^{-3}$. However, as with the estimates above, the standard errors from the SEIR models can become unreasonably large when the variance of the added errors becomes large. Also, as with the SIR models, the $\rr$ estimates and standard errors from the 0th and 1st order polynomial linear model datasets are not very good, as the former is sometimes negative and the latter can be very large. The MSEIR model tends to do fairly well in all situations.

@@

We have assumed for these models that the number of infections follows an exponential curve throughout the entire time frame of the data sets. This assumption certainly does not hold for the data from the 0th and 1st order linear models, and may not hold for some of the other data sets either.

These models are generally not as accurate as the compartment models above. Most of them, except for the initial size model, tend to underestimate $\rr$, with estimates between 0.8 and 1.0. The standard errors are all fairly small, on the order of magnitude from $10^{-1}$ to $10^{-3}$. This is larger than many of the compartment model standard errors. The main advantage that all of these models, except for the maximum likelihood model, seem to be more robust to changing conditions and functional forms of the data compared to the compartment models. The estimates and standard errors for $\rr$ appear to be reasonable in all cases, including for very large added errors and for the 0th and 1st order linear model data.

@@

The branching process model (the only one in this category so far) is not very accurate. In particular, for the SIR datasets, the $\rr$ estimates are less than 1, while the true $\rr$ is greater than 1. The standard errors, though, are relatively small, around the same order of magnitude as the SIR least squares models. Model estimates and standard errors are fairly consistent across the baseline datasets and changing starting compartment sizes. However, increasing the standard deviation of the added errors tends to have a somewhat noticeable impact on model results. The model also does not handle the 0th order polynomial linear data well.

@@

Below, we compare the estimates and standard errors of $\rr$ given by each of the models for a few selected data sets. The forest plots below show the point estimates and 95\% confidence intervals of $\rr$ from each model for a particular data set. The true $\rr$ value is also shown by the vertical red line in the graphs.

All of the figures below show results from data sets that were simulated with added Gaussian errors. The plot on the top left shows the estimates from one of the baseline data sets. For the most part, the $\rr$ values range from approximately 0.8 to 1.4. We note that the range of values includes $\rr$ values of less than 1, meaning that some models predict that a disease is not spreading when it actually is. However, on the whole, the variation of estimates between models is smaller than we originally expected. The confidence intervals are very small relative to the estimates for the most part; exceptions to this include the SIR IPR and the Markov chain model as the exception.

The plot on the top right shows the estimates for the data set where $Y_0$ is increased to 100. The $\rr$ estimates generally seem to be further away from the true $\rr$ value compared to the results from the baseline data sets, and the overall . The confidence intervals are also more often than not larger than those in the baseline. In particular, the standard errors for some of the SIR models increase by several orders of magnitude.

The bottom left plot shows the estimates for the data set where the error added to the generated data is increased. For most of the models, the estimates of $\rr$ do not change very much compared to the baseline data set. The confidence intervals for this data set increase for some of the models, such as the SIR LMA, while they remain approximately the same as the baseline data sets for most of the others.

On the bottom right, we see the $\rr$ estimates for the case when disease prevalence grows linearly, as opposed to exponentially as seen in the data sets generated under SIR conditions, where the disease prevalence grows exponentially. Model performance here varies depending on how well the models the aforementioned linear growth. Some of the models, such as the SIR Markov chain and the IPR models, do not handle this case very well, giving much larger estimates and standard errors than are reasonable. On the other hand, models such as the SIR maximum and final size seem to be more flexible to different trajectories. [The error bars do not show up if they go outside of the x-axis limits - how do we fix this?]
